### 2022-04-07 manuscript.tex
thursday [ci skip]
This build is based on
https://github.com/scherzerthesis/thesis/commit/ca51cc941ab662e7bdc8ce017ea179a06238ba3f.

This commit was created by the following CI build and job:
https://github.com/scherzerthesis/thesis/commit/ca51cc941ab662e7bdc8ce017ea179a06238ba3f/checks
https://github.com/scherzerthesis/thesis/actions/runs/2110624979

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/cfeaf4b301d421f38362c16fb408e8a598ee88be/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/cfeaf4b301d421f38362c16fb408e8a598ee88be}{scherzerthesis/thesis@cfeaf4b}
on April 6, 2022.

@@

@@

\begin{figure}
\hypertarget{fig:09}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/tuba6.png}
\caption{\textbf{Figure 2.6. \emph{SETD2} loss is common in human cancers, an indicator of poor prognosis and frequently occurs with \emph{BRAF\textsuperscript{V600E}} mutations.}(A) Survival analysis with ``Survival'' R package indicates a trend towards poorer overall survival with low \emph{SETD2} mRNA in TGCA Lung adenocarcinoma dataset. (B) \emph{SETD2} mRNA is attenuated in Lung adenocarcinoma compared to normal lung tissue (C) \emph{SETD2} mutations significantly co-occur with BRAF\textsuperscript{V600E} mutations and are present in 40\% of \_BRAF\textsuperscript{V600E}) mutant Lung Adenocarcinoma.}\label{fig:09}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:3.1}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p53_1.png}
\caption{\textbf{Figure 3.1.Cre-mediated wildtype-to-mutant alleles used in study.} (A) Priot to Cre-mediated recombination, normal \emph{Braf} and \emph{Trp53} are expressed from their endogenous loci. (B) After Cre-mediated recombination, the indicated mutant proteins are expressed, in addition to a TdTomato fluorescent protein to mark expression of BRAF\textsuperscript{V600E}.}\label{fig:3.1}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:3.4}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p53_3.png}
\caption{\textbf{Figure 3.4. Mutant-p53 alleles differentially affect the survival of mice due to progressive lung adenocarcinoma.} (A) Kaplan-Meier plot of \emph{Braf\textsuperscript{CAT/+}} mice with p53-mutant alleles. (B) Kaplan-Meier plot of \emph{Braf\textsuperscript{CAT/+}} mice with homozygous p53-mutant alleles.* = \emph{p-value\textless.05}}\label{fig:3.4}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:3.7}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p536.png}
\caption{\textbf{Figure 3.7. Mouse tumors and cell lines that expression P53-R172H and p53-R245W are refractory to MAPK blockade.} (A) Tumor burden analysis of GEMMs treated with MAPKi chow for four weeks. (B) \emph{n vitro} dose response to dabrafenib for 48 hours. (C) Western blots of timecourse of cell lines treated with dabrafenib plus trametinib.}\label{fig:3.7}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:3.13}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p5311.png}
\caption{\textbf{Figure 3.13. PDGFRB is upregulated and expressed in mutant-p53 tumors and cell lines.} (A) Log-normalized counts of \emph{Pdgfrb} RNA reads from p53-mutant cell lines. (B) Western blot analysis of basal and PDGF-BB stimulated cells. (C) IHC of PDGFRB in p53-mutant GEMM tumors.}\label{fig:3.13}
}
\end{figure}

\begin{figure}
\hypertarget{fig:3.14}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p5312.png}
\caption{\textbf{Figure 3.14. PDGF-stimulation or PDGF patway blockade has no effect on cell proliferation. (A) Growth response to PDGF stimulation.} (B) Growth response to Imatinib treatment.}\label{fig:3.14}
}
\end{figure}

@@

\#\#Abstract
malignant lung adenocarcinomas tumors are initiated from a single neoplastic cell. Yet, there is substantial cellular heterogenity in a fully malignant tumor. Since human lung adenocarcinomas are heavily mutated, it is diffuclt to know wheter underlying cellular heterogenity is a consequence of the underlying mutational heterogenity. Mouse models of lung adenocarcinoma accurately represent genomic and histological aspects of human lung adenocarcinoma. However, genetically-engineered mouse model tumor heterogenity remains poorly understood. Here, we profile the single cell transcriptomes of genetically-engineered mouse models that are driven by the BRAF\textsuperscript{V600E} oncoprotein with 5 seperate cooperating genetic events ( \emph{Pi3k\textsuperscript{H1047R}}, \emph{CdkN2a\textsuperscript{fl/fl}}, \emph{Trp53\textsuperscript{fl/fl}}, \emph{Trp53\textsuperscript{R172H/R162H}}, and \emph{Trp53\textsuperscript{R245W/R245W}}). There are stereotypically distinct clusters of cells thorughout all genotypes of BRAF\textsuperscript{V600E}-driven tumors. We also identified known tumor cell identities in both human and mouse lung adenocarcinoma.

@@

The mammalian lung is a vast and intricate labyrinth that is responsible that is responsible for gas exchange between air and body through a meshwork of endothelial cells. The lung organ system begins with a single airway that branches in a geometric and stereotypic pattern(Metzger et al., 2008). These many thousands of branches lead to the business end of gas exchange, the lung alveoli. The alveoli are comprised of two specialized cell types: Alveolar Type 1 (AT1) cells comprise most of the surface area of the lung and are thin squamous cells that mediate oxygen exchange across cell membranes. Alveolar Type 2 (AT2) cells, however, are cuboidal that secrete surfactant proteins that aide in preventing the delicate alveoli from collapsing.

Pathology of the alveoli is a steadfast threat to an organism's life due to the importance of gas exchange. Diseases such as lung adenocarcinoma, pulmonary fibrosis, and emphysema are the most common dysfunctions of the distal lung epithelium(Desai et al., 2014). Lung adenocarcinoma is also the most common type of human lung cancers in both men and women. A large body of literature has suggested that lung adenocarcinomas arise in the alveolar and are descendants of AT2 cells (Ferone et al., 2020; Jackson et al., 2001). Therefore, there is great interest in understanding how AT2 cells develop, maintain cell identity, as well as how they renew themselves during aging or in response to injury (2020; Alanis et al., 2014, 2014; Treutlein et al., 2014).

Recent advances in profiling of single-cell transcriptomes (scRNA Seq) during homeostasis conditions and diseases states have greatly improved our understanding of AT2 cell, and lung adenocarcinoma cell biology(LaFave et al., 2020; Marjanovic et al., 2020; Treutlein et al., 2014).Single-cell profiling of human lung cancers have revealed the drastic intra-tumoral heterogeneity that exists in a single tumor(Gerlinger et al., 2012; Xue et al., 2016). Since transcriptional heterogeneity can be considered a source of therapeutic resistance, there is great interest in understanding the evolution of lung adenocarcinoma and how it changes with response to pathway targeted therapy(Maynard et al., 2020).Since GEMMs are an important pre-clinical model for testing out potential therapeutic strategies for lung adenocarcinoma, it is important to understand single-cell heterogeneity in popular GEMM models.

Although GEMMs are a great way to model human lung adenocarcinomas, there are several drawbacks worth mentioning. Human lung adenocarcinomas typically develop from a single tumor-initiating cell and is a consequence of an accumulation of mutations in certain cell types and a weakened immune system across a lifetime. In GEMMs, tumorigenesis is often initiated in many cells at once and is the consequence of an engineered mutation. In contrast to human lung adenocarcinomas that have a high mutational burden, murine lung adenocarcinomas do not have a high mutational burden and instead display high copy number variation(Westcott et al., 2014). Therefore, we set out to profile the single-cell transcriptomes of various BRAF\textsuperscript{V600E}-driven lung cancer GEMMs in order to determine genotype-specific heterogeneity and response to pathway-targeted therapy.

@@

\textbf{ScRNA Sequencing of benign and malignant BRAF\textsuperscript{V600E} mouse lung tumors}
We profiled the single-cell transcriptome of six different BRAF\textsuperscript{V600E} mouse lung tumor models including BRAF\textsuperscript{V600E} on its own, BRAF\textsuperscript{V600E} plus either mutant \emph{Pik3ca\textsuperscript{H1047R/+}}, \emph{Cdkn2a\textsuperscript{fl/fl}}, \emph{Trp53\textsuperscript{fl/fl}}, \emph{Trp53\textsuperscript{R172H/R172H}}, or \emph{Trp53\textsuperscript{R245W/R245W}}. We also treated each genotype with dabrafenib (100mg/kg) plus trametinib(2mg/kg) once daily for four days prior to tissue harvesting and 10X genomics library preperation.Importantly, we kept dabrafenib and trametinib throughout all stages of processing to prevent altered transcription in response to drug cessation. Using fluorescently-activated cell sorting (FACS), we purified tumor cells based on Td-Tomato expression 8 weeks after tumor initiation using adenoviral Spc-Cre. In sum, we were able to identify 71,110 cells that passed standard 10X genomics and Seuratbased quality control and filtering (Figure 4.1).

Although we sought to purify tumor cells based on TdTomato-positivity, we did unintentionally collect and analyze stromal cells, such as immune, mesenchymal, and endothelial cell, which are also present in the murine distal lung epithelium (Figure 4.2). We next determined tumor cells by previously identified tumor cell transcriptional programs in KRAS(G12D) mouse lung cancer(Marjanovic et al., 2020). We succesfully identified AT2-like, AT1/AT2 bipotent progenitor, Cycling, mixed, and Hepatic-Gastric cellular programs in our dataset(Figure 4.3). We used these programs to purify tumor cells for downstream differential gene expression analysis. Reclustering of tumor cell revealed significant tumor cell heterogeneity(Figure 4.4). Next we sought to determine cluster specific markers using DESeq2 analysis (Figure 4.4B). Thus, we have successfully documented the tumor heterogenity in various models of BRAF\textsuperscript{V600E} lung tumorigenesis.

We next sought to determine the effect of dabrafenib plus trametinib on transcriptional heterogeneity (Figure 4.5). There is a MAPK specific gene expression program that is decreased in response to MAPKi(Figure 4.5A). Clearly, all genotypes respond similary to MAPK-pathway inhibition(Figure 4.5B). Therefore, we hypothesize that although there is transcriptional heterogeneity, there is a uniform response to MAPK-blockade.

Since there is great interest in p53-mediated tumor supression, we wondered if we could detect a p53-gene signature in our data set. Indeed, we did detect the reduction of transcripts that are often associated with p53 transcriptional activation in reponse to p53-mutation, such as downregulation of \emph{Cdkn1a}, \emph{Zmat3}, and \emph{CCng1}(Figure 4.6A). Furthermore, P53-target gene expression is maintained only in p53 wildtpype samples, including \emph{Cdkn2a} loss (Figure 4.6B). Therefore we have successful identified a p53-gene signature in \_BRAF\textsuperscript{V600E} expressing lung tumors using scRNA sequncing.

We were also interested in genotype-specific differences. therefore, we performed DESEQ2 between vehicle-treated mice of each genotype (Figure 4.7). Although we did find differentially expressed genes between genotypes, there appears to be significant batch effects, as samples cluster based on their 10X genomic chip runs. Therefore, more careful analysis is needed in order to determine suble genotype-specific gene expression changes. For instance, there are a variety of batch-correcting pipelines we could utilize to negate batch effects.

@@

Here, we used scRNA-seq to determine transcriptional heterogenity of BRAF\textsuperscript{V600E} driven lung cancer and how it changes in response to concomminant genetic alterations. We modeled lung adenocarcinoma that mimick human oncogenic transformation where MAPK pathway activation drives tumor formation, and inactivation of the p53 pathway allows for tumor progression(Campbell et al., 2016). We also determined how cell state changes in response to MAPK blockade, a common theraputic option for patients with BRAF\textsuperscript{V600E}-mutant lung adenocarcinomas.

We determined that there is transcriptional heterogenity and that it is reporducible across genotypes. We also confirmed previousl published cell states in similar models, such as KRAS G12D(Marjanovic et al., 2020). Unfortunely, we did not observe robust dedifferentiation signals such as loss of NKX2-1 and adoption of a gastric cell state. However, we did collect and identify highly-pastic cell states, such as ITGA2+ and CLDN4+ cells. Our failure to collect and analyze pre-metastatic cell states such as EMT-like cells, is likely because we collected cells only after 8-weeks, which is likely not long enough for these cell states to evolve during lung tumorigenesis. Therefore, adoption of highly malignant cell states, although might require certain genomic alterations like p53 loss, is an outcome of tumor evolution. One hypothesis is that p53 loss causes genomic instability and thus copy number gain, but similar studies failed to detect robust changes in copy number variation(Marjanovic et al., 2020).

We also failed to detect robust differences between genotypes that are known to cause malignant transformation, such as p53 loss, p53 missense mutation, CDKN2A loss, or mutant PI3K\textsuperscript{H1047R} (Dankort et al., 2007; Trejo et al., 2013; van Veen et al., 2019). This is likely a result from a combination of factors, such as time of collection and library capture/sequencing depth. One straightforward hypothesis is that we did not allow for full transcriptomic changes to take effect that is imparted by tumor suppressor or oncogene mutation. Therefore, a timecourse analysis that spans early timepoints (4-6 weeks) and late timepoints (\textgreater16 weeks) will likely reveal more informative transcriptional heterogeneity. Indeed, our previous studies have indicated that there is no tumor growth advantage at early time points as late as 6 weeks post tumor initiation. One other possibility that could explain our lack of genotype-specific gene expression programs is that we did not capture enough unique transcripts in order to reveal the full extent of tumor heterogeneity. This is likely because there are known limitations to 10X genomic library preperation. Specifically, there is gene-dropout effects where only the most highly expressed genes are captured by the droplet-based technique. Indeed, 10X genomics technologies have been shown not to detect genes with lower expression(Wang et al., 2021). However, 10X genomics allows you to profile more cells than similiar scRNA-seq techniques. Genes with lower expression can be just as important as highly expressed genes for a cell's identity. Actually, transcription factors that determine cell states have been shown to have modest expression{[}Wang et al. (2021); (\textbf{pmid?});30205475{]}.
Therefore, we hypothesis that using plate-based scRNA-seq approaches, such as Smart-Seq2, would allow for more gene capture. Thus, Smart-seq2 will enable us to more precisely determine transcirptional heterogenity and genotype-specific gene expression changes.

Although our analysis was lacking with regards to transcriptional differences between genotypes, we did detect a robust transcriptional change with MAPK blockade (figure4.5). Since human lung cancer patients treated with MAPK-pathway inhibitors exhibit a heterogeneity in response, one might hypothesize that each cell responds differently to MAPK-pathway inhibition. In our analysis, we determine that there is generally a homogenous response to MAPK inhibition. One straightforward explanation is that our short term treatment (four days) did not allow for cells to respond to pathway blockade. Perhaps there is an intial response to MAPK-pathway inhibition that is dampened over time that is dependant on transcriptional heterogeneity.

In conclusion, we used 10X genomics to determine the transcriptional heterogeneity of BRAF\textsuperscript{V600E} lung tumors with a five seperate cooperating alterations and treatments. We did identify transciprtional heterogeneity that is stereotypic and reproducible acorss genotypes. However, we failed to identify robust differences between genotypes. Our works highlights the importance of capturing full transcriptomes across multiple timepoints in order to accurately determine cell states throughout BRAF\textsuperscript{V600E}-driven lung tumor evolution.

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/scrna_1.png}
\caption{\textbf{Figure 4.1. 71,110 cells collected from different genotypes.}}\label{fig:04}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/scrna_6.png}
\caption{\textbf{Figure 4.7. Genotpye-specific gene expression changes.}}\label{fig:04}
}
\end{figure}

\hypertarget{materials-and-methods}{%
\subsection{Materials and Methods}\label{materials-and-methods}}

>>>>

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/ca51cc941ab662e7bdc8ce017ea179a06238ba3f/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/ca51cc941ab662e7bdc8ce017ea179a06238ba3f}{scherzerthesis/thesis@ca51cc9}
on April 7, 2022.

@@

\#\#Methods

\textbf{Animal Husbandry} All mice were housed in an environmentally controlled room, and all animal care and experimental procedures were approved by (and in accordance with) the Institutional Animal Care and Use Committee Office of the Huntsman Cancer Institute at the University of Utah. Genetically engineered mouse breeding and genotyping was conducted as previously described (van Veen et al., 2019). The \emph{BRAF\textsuperscript{CAT}} and mice were previously described (Dankort et al., 2007; Jonkers et al., 2001; Shai et al., 2015; van Veen et al., 2019). \emph{H11b\textsuperscript{LSL-Cas9}} mice were previously described(Chiou et al., 2015) All viruses were administered in a Biosafety Level 2+ room, as is regulated by the Institutional Biosafety Committee Guidelines. Lentivirus (Stanford University) was delivered through intratracheal intubation.

\textbf{Histology and Quantification of lung tumor burden}
Lungs were perfused with neutral-buffered formalin, removed, and kept in neutral-buffered formalin for 16-24 hours and then stored in 70\% ethanol prior to embedding. 5 micrometer sections were cut using the HCI histology core. sections were then stained with hematoxylin and eosin (H\&E). Stained slides were scanned using the 3D-histech Panoramic MIDI slide scanner. Tumor size was quantifie using 3D Histech QuantCenter. Immunohistochemistry for H3K26me\textsuperscript{3} was performed as previously described at a concentration of 1:1000 with rabbit mAb \#4909 from Cell Signaling Technologies. (Mollaoglu et al., 2017; van Veen et al., 2019).

\textbf{Human lung cancer data} cBio Portal and Project Genie were used to calculate co-ocurring mutations with \emph{BRAF\textsuperscript{V600}} mutations. KMplotter was used for survival analysis and TNMplot was used to calculate \emph{SETD2} expression in tumor tissue vesus normal lung tissue(Bartha and Győrffy, 2021; Lánczky and Győrffy, 2021).

@@

\begin{figure}
\hypertarget{fig:09}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/tuba6.png}
\caption{\textbf{Figure 2.6. \emph{SETD2} loss is common in human cancers, an indicator of poor prognosis and frequently occurs with \emph{BRAF\textsuperscript{V600E}} mutations.}(A) Survival analysis with ``KMplotter'''' indicates a trend towards poorer overall survival with low \emph{SETD2} mRNA in TGCA Lung adenocarcinoma dataset. (B) \emph{SETD2} mRNA is attenuated in Lung adenocarcinoma compared to normal lung tissue (C) \emph{SETD2} mutations significantly co-occur with BRAF\textsuperscript{V600E} mutations and are present in 40\% of \_BRAF\textsuperscript{V600E}) mutant Lung Adenocarcinoma.}\label{fig:09}
}
\end{figure}

@@

\#\#Methods

\begin{figure}
\hypertarget{fig:3.1}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p53_1.png}
\caption{\textbf{Figure 3.1. Schematic of \emph{Braf} and \emph{Trp53} alleles} (A) Priot to Cre-mediated recombination, normal \emph{Braf} and \emph{Trp53} are expressed from their endogenous loci. (B) After Cre-mediated recombination, the indicated mutant proteins are expressed, in addition to a TdTomato fluorescent protein to mark expression of BRAF\textsuperscript{V600E}-positive cells.}\label{fig:3.1}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:3.4}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p53_3.png}
\caption{\textbf{Figure 3.4. Mutant-p53 alleles differentially affect the survival of mice due to progressive lung adenocarcinoma.} (A) Kaplan-Meier plot of \emph{Braf\textsuperscript{CAT/+}} mice with heterozygous p53-mutant alleles. (B) Kaplan-Meier plot of \emph{Braf\textsuperscript{CAT/+}} mice with homozygous p53-mutant alleles.* = \emph{p-value\textless.05}}\label{fig:3.4}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:3.7}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p536.png}
\caption{\textbf{Figure 3.7. Mouse tumors and cell lines that expression P53-R172H and p53-R245W are refractory to MAPK blockade.} (A) Tumor burden analysis of GEMMs treated with MAPKi chow for four weeks. (B) \emph{in vitro} dose response to dabrafenib for 48 hours. (C) Western blots of time-course of cell lines treated with dabrafenib plus trametinib.}\label{fig:3.7}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:3.13}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/pdgfr_1.png}
\caption{\textbf{Figure 3.13. PDGFRB is upregulated and expressed in mutant-p53 tumors and cell lines.} (A) Log-normalized counts of \emph{Pdgfrb} RNA reads from p53-mutant cell lines. (B) Western blot analysis of basal and PDGF-BB stimulated cells. (C) IHC of PDGFRB in p53-mutant GEMM tumors.}\label{fig:3.13}
}
\end{figure}

\begin{figure}
\hypertarget{fig:3.14}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/pdgfr_2.png}
\caption{\textbf{Figure 3.14. PDGF-stimulation or PDGF patway blockade has no effect on cell proliferation. (A) Growth response to PDGF stimulation.} (B) Growth response to Imatinib treatment.}\label{fig:3.14}
}
\end{figure}

@@

\#\#Abstract
Malignant lung adenocarcinomas tumors are initiated from a single neoplastic cell. Nevertheless, there is substantial cellular heterogeneity in a fully malignant tumor. Since human lung adenocarcinomas are heavily mutated, it is difficult to know whether underlying cellular heterogeneity is a consequence of the underlying mutational heterogeneity. Mouse models of lung adenocarcinoma accurately represent genomic and histological aspects of human lung adenocarcinoma. However, genetically-engineered mouse model tumor heterogeneity remains poorly understood. Here, we profile the single-cell transcriptomes of genetically-engineered mouse models driven by the BRAF\textsuperscript{V600E} oncoprotein with five separate cooperating genetic events ( \emph{Pi3k\textsuperscript{H1047R}}, \emph{CdkN2a\textsuperscript{fl/fl}}, \emph{Trp53\textsuperscript{fl/fl}}, \emph{Trp53\textsuperscript{R172H/R162H}}, and \emph{Trp53\textsuperscript{R245W/R245W}}). There are stereotypically distinct clusters of cells throughout all genotypes of BRAF\textsuperscript{V600E}-driven tumors. We also identified known tumor cell identities in human and mouse lung adenocarcinoma.

@@

The mammalian lung is a vast and intricate labyrinth responsible for gas exchange between air and body through a meshwork of endothelial cells. The lung organ system begins with a single airway that branches in a geometric and stereotypic pattern(Metzger et al., 2008). These thousands of branches lead to the business end of gas exchange, the lung alveoli. The alveoli comprise two specialized cell types: Alveolar Type 1 (AT1) cells comprise most of the lung's surface area and are thin squamous cells that mediate oxygen exchange across cell membranes. However, alveolar Type 2 (AT2) cells are cuboidal that secrete surfactant proteins that aid in preventing the delicate alveoli from collapsing.

Pathology of the alveoli is a steadfast threat to an organism's life due to the importance of gas exchange. Diseases such as lung adenocarcinoma, pulmonary fibrosis, and emphysema are the most common dysfunctions of the distal lung epithelium(Desai et al., 2014). Lung adenocarcinoma is also the most common type of human lung cancer in men and women. A large body of literature has suggested that lung adenocarcinomas arise in the alveolar and are descendants of AT2 cells (Ferone et al., 2020; Jackson et al., 2001). Therefore, there is great interest in understanding how AT2 cells develop and maintain cell identity and how they renew themselves during aging or in response to injury (2020; Alanis et al., 2014, 2014; Treutlein et al., 2014).

Recent advances in profiling single-cell transcriptomes (scRNA Seq) during homeostasis conditions and disease states have greatly improved our understanding of AT2 cell and lung adenocarcinoma cell biology(LaFave et al., 2020; Marjanovic et al., 2020; Treutlein et al., 2014). Single-cell profiling of human lung cancers has revealed the extreme intra-tumoral heterogeneity in a single tumor(Gerlinger et al., 2012; Xue et al., 2016). Since transcriptional heterogeneity can be considered a source of therapeutic resistance, there is great interest in understanding the evolution of lung adenocarcinoma and how it changes with response to pathway targeted therapy(Maynard et al., 2020). Since GEMMs are an important pre-clinical model for testing potential therapeutic strategies for lung adenocarcinoma, it is essential to understand single-cell heterogeneity in popular GEMM models.

Although GEMMs are a great way to model human lung adenocarcinomas, several drawbacks are worth mentioning. Human lung adenocarcinomas typically develop from a single tumor-initiating cell. They result from an accumulation of mutations in specific cell types and a weakened immune system across a lifetime. In GEMMs, tumorigenesis is often initiated in many cells at once and is the consequence of an engineered mutation. In contrast to human lung adenocarcinomas, which have a high mutational burden, murine lung adenocarcinomas do not have a high mutational burden and instead display high copy number variation(Westcott et al., 2014). Therefore, we set out to profile the single-cell transcriptomes of various BRAF\textsuperscript{V600E}-driven lung cancer GEMMs to determine genotype-specific heterogeneity and response to pathway-targeted therapy.

\#\#Methods

@@

We profiled the single-cell transcriptome of six different BRAF\textsuperscript{V600E} mouse lung tumor models, including BRAF\textsuperscript{V600E} on its own, BRAF\textsuperscript{V600E} plus either mutant \emph{Pik3ca\textsuperscript{H1047R/+}}, \emph{Cdkn2a\textsuperscript{fl/fl}}, \emph{Trp53\textsuperscript{fl/fl}}, \emph{Trp53\textsuperscript{R172H/R172H}}, or \emph{Trp53\textsuperscript{R245W/R245W}}. We also treated each genotype with dabrafenib (100mg/kg) plus trametinib(2mg/kg) once daily for four days before tissue harvesting and 10X genomics library preparation. Importantly, we kept dabrafenib and trametinib throughout all processing stages to prevent altered transcription in response to drug cessation. Using fluorescently-activated cell sorting (FACS), we purified tumor cells based on Td-Tomato expression eight weeks after tumor initiation using adenoviral Spc-Cre. We were able to identify 71,110 cells that passed standard 10X genomics and Seuratbased quality control and filtering (Figure 4.1).

Although we sought to purify tumor cells based on TdTomato-positivity, we unintentionally collected and analyzed stromal cells, such as immune, mesenchymal, and endothelial cells, which are also present in the murine distal lung epithelium (Figure 4.2). We next determined tumor cells by previously identified tumor cell transcriptional programs in KRAS(G12D) mouse lung cancer(Marjanovic et al., 2020). In our dataset, we successfully identified AT2-like, AT1/AT2 bipotent progenitor, Cycling, mixed, and Hepatic-Gastric cellular programs (Figure 4.3). We used these programs to purify tumor cells for downstream differential gene expression analysis. Reclustering of tumor cells revealed significant tumor cell heterogeneity(Figure 4.4). Next, we sought to determine cluster-specific markers using DESeq2 analysis (Figure 4.4B). Thus, we have successfully documented the tumor heterogeneity in various models of BRAF\textsuperscript{V600E} lung tumorigenesis.

We next sought to determine the effect of dabrafenib plus trametinib on transcriptional heterogeneity (Figure 4.5). There is a MAPK specific gene expression program is decreased in response to MAPKi(Figure 4.5A). All genotypes respond similarly to MAPK-pathway inhibition(Figure 4.5B). Therefore, we hypothesize that although transcriptional heterogeneity exists, there is a uniform response to MAPK-blockade.

Since there is great interest in p53-mediated tumor suppression, we wondered if we could detect a p53-gene signature in our data set. Indeed, we detected the reduction of transcripts often associated with p53 transcriptional activation in response to p53-mutation, such as downregulation of \emph{Cdkn1a}, \emph{Zmat3}, and \emph{CCng1}(Figure 4.6A). Furthermore, P53-target gene expression is maintained only in p53 wild-type samples, including \emph{Cdkn2a} loss (Figure 4.6B). Therefore we have successfully identified a p53-gene signature in \_BRAF\textsuperscript{V600E} expressing lung tumors using scRNA sequencing.

We were also interested in genotype-specific differences. Therefore, we performed DESEQ2 between vehicle-treated mice of each genotype (Figure 4.7). Although we did find differentially expressed genes between genotypes, there appear to be significant batch effects, as samples cluster based on their 10X genomic chip runs. Therefore, more careful analysis is needed to determine subtle genotype-specific gene expression changes. For instance, we could utilize a variety of batch-correcting pipelines to negate batch effects.

@@

Here, we used scRNA-seq to determine transcriptional heterogeneity of BRAF\textsuperscript{V600E} driven lung cancer and how it changes in response to concomitant genetic alterations. We modeled lung adenocarcinoma that mimick human oncogenic transformation where MAPK pathway activation drives tumor formation, and inactivation of the p53 pathway allows for tumor progression(Campbell et al., 2016). We also determined how cell state changes in response to MAPK blockade, a standard therapeutic option for patients with BRAF\textsuperscript{V600E}-mutant lung adenocarcinomas.

We determined that there is transcriptional heterogeneity and that it is reproducible across genotypes. We also confirmed previously published cell states in similar models, such as KRAS G12D(Marjanovic et al., 2020). Unfortunately, we did not observe robust dedifferentiation signals such as loss of NKX2-1 and a gastric cell state adoption. However, we did collect and identify highly-plastic cell states, such as ITGA2+ and CLDN4+ cells. Our failure to collect and analyze pre-metastatic cell states such as EMT-like cells is likely because we collected cells only after 8-weeks, which is likely, not long enough for these cell states to evolve during lung tumorigenesis. Therefore, although it might require specific genomic alterations like p53 loss, the adoption of highly malignant cell states is an outcome of tumor evolution. One hypothesis is that p53 loss causes genomic instability and thus copy number gain, but similar studies failed to detect robust changes in copy number variation(Marjanovic et al., 2020).

We also failed to detect robust differences between genotypes known to cause malignant transformation, such as p53 loss, p53 missense mutation, CDKN2A loss, or mutant PI3K\textsuperscript{H1047R} (Dankort et al., 2007; Trejo et al., 2013; van Veen et al., 2019). This is likely due to a combination of factors, such as time of collection and library capture/sequencing depth. One straightforward hypothesis is that we did not allow complete transcriptomic changes to take effect imparted by tumor suppressor or oncogene mutation. Therefore, a time-course analysis that spans early time points (4-6 weeks) and late time points (\textgreater16 weeks) will likely reveal more informative transcriptional heterogeneity. Indeed, our previous studies have indicated no tumor growth advantage at early time points as late as six weeks post tumor initiation. One other possibility that could explain our lack of genotype-specific gene expression programs is that we did not capture enough unique transcripts to reveal the full extent of tumor heterogeneity. This is likely because there are known limitations to 10X genomic library preparation. Specifically, there are gene-dropout effects where the droplet-based technique captures only the most highly expressed genes. Indeed, 10X genomics technologies have been shown not to detect genes with lower expression(Wang et al., 2021). However, 10X genomics allows for profiling more cells than similar scRNA-seq techniques. Genes with lower expression can be as important as highly expressed genes for a cell's identity. Transcription factors that determine cell states have been shown to have modest expression{[}Wang et al. (2021); (\textbf{pmid?});30205475{]}.
Therefore, we hypothesize that using plate-based scRNA-seq approaches, such as Smart-Seq2, would allow more gene capture. Thus, Smart-seq2 will enable us to determine transcriptional heterogeneity and genotype-specific gene expression changes more precisely.

Although our analysis was lacking regarding transcriptional differences between genotypes, we did detect a robust transcriptional change with MAPK blockade (figure4.5). Since human lung cancer patients treated with MAPK-pathway inhibitors exhibit heterogeneity, one might hypothesize that each cell responds differently to MAPK-pathway inhibition. Our analysis determines that there is generally a homogenous response to MAPK inhibition. One straightforward explanation is that our short-term treatment (four days) did not allow cells to respond to pathway blockade. Perhaps there is an initial response to MAPK-pathway inhibition that is dampened over time dependent on transcriptional heterogeneity.

In conclusion, we used 10X genomics to determine the transcriptional heterogeneity of BRAF\textsuperscript{V600E} lung tumors with five separate cooperating alterations and treatments. We did identify transcriptional heterogeneity that is stereotypic and reproducible across genotypes. However, we failed to identify robust differences between genotypes. Our work highlights the importance of accurately capturing complete transcriptomes across multiple time points to accurately determine cell states throughout BRAF\textsuperscript{V600E}-driven lung tumor evolution.

\#\#Methods
\textbf{Animal Husbandry} All mice were housed in an environmentally controlled room, and all animal care and experimental procedures were approved by (and in accordance with) the Institutional Animal Care and Use Committee Office of the Huntsman Cancer Institute at the University of Utah. Genetically engineered mouse breeding and genotyping was conducted as previously described (van Veen et al., 2019). The \emph{BRAF\textsuperscript{CAT}}, \emph{Trp53\textsuperscript{fl/fl}}, \emph{Cdkn2a\textsuperscript{fl/fl}}, and \emph{Pik3ca\textsuperscript{H1047R/+}} mice were previously described (Dankort et al., 2007; Jonkers et al., 2001; Shai et al., 2015; van Veen et al., 2019). All viruses were administered in a Biosafety Level 2+ room, as is regulated by the Institutional Biosafety Committee Guidelines. Adeno-SPC-CRE virus (University of Iowa) was delivered through nasal instillation. The \emph{Trp53\textsuperscript{wmR172H}} and \emph{Trp53\textsuperscript{wmR245W}} mice and respective genotyping protocols were gifted by Gigi Lozano and her laboratory and previously published (Kim et al., 2021; Zhang et al., 2018).

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/scrna_1.png}
\caption{\textbf{Figure 4.1. 71,110 cells collected from six different genotypes.}}\label{fig:04}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/scrna_6.png}
\caption{\textbf{Figure 4.7. Heatmap of top Genotpye-specific gene expression changes.}}\label{fig:04}
}
\end{figure}
